---
title: "W20 Bio201 Final Project"
author: "Akshaya Ravikumar"
date: "4/21/2020"
output:
  pdf_document: default
  html_document: default
---

#Research Question
What Might be the Key Correlational Explanatory Variable(s) for Decrease in Butyrate Concentration after Supplement Consumption?

#H0
There are no similarities in regards to community composition, taxa, and/or other descriptive data in those with negative/no butyrate response.

#Ha 
There are significant similarities in regards to community composition, taxa, and/or other descriptive data in those with negative/no butyrate response.

#Git Repository: 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMBio201/Final Project/Projects/")
tinytex::install_tinytex()
```

# Load Packages
```{r include=FALSE}
library(vegan)
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(phyloseq)
set.seed(7)
source("miseqR.R")
```

# Load Data
Import the sample measurements and data.  
```{r}
# name: sample_df

sample_indiv_df <- read_delim("sample_measurements_indv.txt", 
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE,
                            col_types = cols( .default = col_character(),
                                               Sample_number = col_double(),
                                               Final_weight = col_double(),
                                               Acetate_mM = col_double(),
                                               Acetate_mmol_kg = col_double(),
                                               Butyrate_mM = col_double(),
                                               Butyrate_mmol_kg = col_double(),
                                               Propionate_mM = col_double(),
                                               Propionate_mmol_kg = col_double(),
                                               pH = col_logical(),
                                               Status = col_logical(),
                                              Bristol_score = col_logical(),
                                              Bristol_numeric = col_logical())) %>%
  # make all column names lower case
  rename_all(tolower) 

#Weekly has weekly averages per participant, use weekly for now
sample_weekly_df <- read_delim("sample_measurements_wkly.txt", 
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE) %>%
  # make all column names lower case
  rename_all(tolower) %>%
  mutate(id = paste(participant_id, study_week, sep = "_")) 
```

##Import the shared table.
```{r}
#name: shared_m
shared_m <- read_delim("~/Documents/UMBio201/Projects/shared_table_wkly.txt",
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list()) %>%
             # remove duplicate sample ids
              distinct(., ID, .keep_all = TRUE)%>%
    select(ID, starts_with("Otu")) %>%
  # drop control samples from sequencing
  filter(str_detect(ID, "^U")) %>%
  # remove duplicate sample ids
  distinct(ID, .keep_all = TRUE) %>% 
  # sample IDs need to be made into row names
  column_to_rownames(var = "ID") %>% 
  # convert data frame to matrix object
  as.matrix() %>% 
  # convert to phyloseq object 
  otu_table(., taxa_are_rows = FALSE) 


             


```


Import the taxonomy table.
```{r}
#name: taxa_m

taxa_m <- read_delim("taxonomy_table.txt",
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list()) %>%
               # sequence variants (OTUs) need to be made into row names 
  column_to_rownames(var = "OTUs") %>% 
  as.matrix() %>%
  # convert to phyloseq object 
  tax_table() 
 


```

##Create a phyloseq object 
```{r}
# name the result: physq_enterotype_number (e.g., physq_4)

sample_weekly_m <- sample_weekly_df %>%
  # make all column names lower case
  rename_all(tolower) %>% 
  # remove duplicate sample ids --\
  distinct(., participant_id, .keep_all = TRUE) %>%  
  # sample IDs need to be made into row names
  column_to_rownames(var = "id") %>% 
  # specify type of phyloseq object
  sample_data() 
  



physq_final1 <- phyloseq(shared_m, taxa_m, sample_weekly_m) %>% 
  # subset for consent and compliance
  subset_samples(., quantity_compliant != "none") %>%
  # subset for weeks of interest
  subset_samples(., study_week == "week1" | study_week == "week3") %>%
  # subset for potato supplements 
  subset_samples(., supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT") 
```

#Calculate Butyrate Delta and find Positives and Negatives
```{r}

butyrate_delta <- sample_weekly_df %>%
  # keep only columns of interest
  select(participant_id, semester, supplement_consumed, 
         frequency, study_week, butyrate_mean)%>% 
  # keep only weeks of interest
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "LOODAT" | supplement_consumed == "BRMPS") %>% 
  # widen data frame
  spread(study_week, butyrate_mean)  %>%
  # drop NAs from select columns
  drop_na(week1, week3) %>% 
  # calculate difference
  mutate(delta_but = week3 - week1) %>%  
  # drop extra columns
  select(-week3, -week1) %>%
  add_column(butyrate_status = "-")

butyrate_delta$butyrate_status[butyrate_delta$delta_but > 0] <- "positive"
butyrate_delta$butyrate_status[butyrate_delta$delta_but < 0] <- "negative"
butyrate_delta$butyrate_status[butyrate_delta$delta_but == 0] <- "no change"
```

##Summary Statistics for Butyrate Delta Data Frame
```{r}

summary(butyrate_delta$delta_but)
```


##Create data frames derived from 
```{r}
#Create a df with only negative or no change 

butyrate_nres <- butyrate_delta %>%
  filter(butyrate_status != "positive")

butyrate_negdf <- merge(butyrate_nres, sample_weekly_df, by = "participant_id", all.x=TRUE) %>%
  filter(study_week == "week1" | study_week == "week3")

butyrate_df <- merge(butyrate_delta, sample_weekly_df, by = "participant_id", all.x=TRUE) %>%
    filter(study_week == "week1" | study_week == "week3") %>%
    select(-ph_median, -ph_mean, -bristol_median, -bristol_mean, -blood_glucose_median, -blood_glucose_mean,
           -acetate_median, -acetate_mean, -propionate_median, -propionate_mean, -butyrate_median, -id) %>%
    spread(study_week, butyrate_mean)  %>%
  # drop NAs from select columns
  drop_na(week1, week3) %>% 
  gather(study_week, butyrate_mean, week1, week3) %>%
  mutate(id = paste(participant_id, study_week, sep = "_"))



```

There are 102 instances where after potato starch consumption, the butyrate concentration decreases.

Therefore, the next step is to identify commonality among these individuals.

# Butyrate Decrease Test

Is there a significant decrease amongst the negative response individuals?
```{r}

but_wk1_df <- butyrate_negdf %>%
  filter(study_week == "week1") %>%
  rename(but_wk1 = "butyrate_mean" ) %>%
  select(-study_week)

but_wk3_df <- butyrate_negdf %>%
  filter(study_week == "week3") %>%
  rename(but_wk3 = "butyrate_mean") %>%
  select(-study_week)

but_plot <- butyrate_negdf %>%
  ggplot(aes(x = study_week, y = butyrate_mean )) +
  geom_violin(aes(color = study_week)) + geom_jitter(aes(color = study_week)) +
  facet_grid(~frequency.x) +
  xlab("Study Week") +
  ylab("Butryate Mean")+
  theme(legend.position = "none")
but_plot


summarise(but_wk1_df, sample_size = n())

shapiro.test(but_wk1_df$but_wk1) #p-value = 0.3916

ggplot(but_wk1_df, aes(x = but_wk1)) + geom_histogram()

qqnorm(but_wk1_df$but_wk1); qqline(but_wk1_df$but_wk1) 

shapiro.test(but_wk3_df$but_wk3) #p-value = 0.1148

ggplot(but_wk3_df, aes(x = but_wk3)) + geom_histogram()

qqnorm(but_wk3_df$but_wk3); qqline(but_wk3_df$but_wk3) 



var.test(x = but_wk1_df$but_wk1, 
         y = but_wk3_df$but_wk3, 
         alternative = "two.sided") 
#p-value = 0.004636, variances not equal
    
    
    ##But
t.test(x = but_wk1_df$but_wk1, 
       y = but_wk3_df$but_wk3,
       paired = TRUE, 
       var.equal = FALSE, 
       alternative = "greater")
#p-value: <2.2e-16, very significant decrease

#set alternative to greater, since the first week should have a greater mean butyrate than the second


#save plot
save_plot(but_plot, 
          nrow = 2 , ncol = 2, 
          filename = "~/Documents/UMBio201/Final Project/Projects/but_plot.pdf")





```
Conclusion: t-test p-value of <2.2e-16 indicates that the difference in butyrate concentration for negative respondents.

# Butyrate Delta Phyloseq

```{r}


butyrate_m <- butyrate_df %>%
  # make all column names lower case
  rename_all(tolower) %>% 
  # remove duplicate sample ids 
  # sample IDs need to be made into row names
  column_to_rownames(var = "id") %>% 
  # specify type of phyloseq object
  sample_data() 

physq_but <- phyloseq(shared_m, taxa_m, butyrate_m) %>% 
  # subset for consent and compliance
  subset_samples(., quantity_compliant != "none") %>%
  # subset for weeks of interest
  subset_samples(., study_week == "week1" | study_week == "week3") %>%
  # subset for potato supplements 
  subset_samples(., supplement_consumed.y == "BRMPS" | supplement_consumed.y == "LOODAT") 


```


# Relative Abundance
The code below filters for participants who had negative status, then plots the relative abundance of each phyla found in each participant. 
```{r}
physq_phyla <- physq_but %>% 
  # tax glom groups together taxa with the same name
  tax_glom("Phylum") %>%
  # convert counts to relative abundance 
  transform_sample_counts(function(x) {x/sum(x)} ) %>%
  # convert phyloseq object (matrix) to data frame 
  psmelt() %>%
  # sort alphabetically 
  arrange(Phylum) %>%
  # Calculate weekly mean of relative abundance
  group_by(participant_id, study_week, semester.y, 
           frequency.y, butyrate_status, Phylum, OTU) %>%
  summarise(relative_abundance = mean(Abundance, na.rm = TRUE)) %>%
  # remove low abundance taxa
  filter(relative_abundance > 0.001)
physq_phyla
```

##Plot for Negative Status Week 1
```{r}
negbut_plotwk1 <- physq_phyla %>%
  # filter for supplement and study week 
  filter(butyrate_status == "negative",
         study_week == "week1") %>% 
  # set parameters for plot
  ggplot(aes(x = participant_id, y = relative_abundance,
             fill = Phylum,
             study_week, frequency)) + 
  # layer for stacked bar plot 
  geom_bar(stat = "identity") +
  # clean up x-axis 
  theme(axis.title.x = element_blank(), #remove x-axis label 
        axis.text.x =  element_text(angle = 90, #rotate tick labels 
                                    vjust = 0.5,
                                    hjust = 0.5,
                                    size = 8)) +
  # clean up y-axis
  ylab("Relative Abundance\n(% total sequences)") +
  ggtitle("Week 1 - Negative Butyrate Status") +
  scale_fill_manual(name = "Phylum",
                  #these are all the phylum categories
                  breaks = c("Actinobacteria", "Bacteria_unclassified", 
                             "Bacteroidetes", "Elusimicrobia", 
                             "Firmicutes", "Fusobacteria", 
                             "Lentisphaerae", "Proteobacteria", 
                             "Sprichaetes", "Synergistetes", 
                             "Verrucomicrobia"), 
                  #these are updated names for the phyla
                  labels = c("Actinobacteria", "Unclassified", 
                             "Bacteroidetes", "Elusimicrobia", 
                             "Firmicutes", "Fusobacteria", 
                             "Lentisphaerae", "Proteobacteria", 
                             "Sprichaetes", "Synergistetes", 
                             "Verrucomicrobia"), 
                  #colors for each category
                  values = c("#4bb092", "#618bcb",
                             "#b95d6a", "#79853d",
                             "#c783c1", "#bd7a43",
                             "#c9aa3c", "#d4447a",
                             "#7663cb", "#67b34a",
                             "#d35238")  )
negbut_plotwk1
```

##Plot for Negative Status Week 3
```{r}
negbut_plotwk3 <- physq_phyla %>%
  # filter for supplement and study week 
  filter(butyrate_status == "negative",
         study_week == "week3") %>% 
  # set parameters for plot
  ggplot(aes(x = participant_id, y = relative_abundance,
             fill = Phylum,
             study_week, frequency)) + 
  # layer for stacked bar plot 
  geom_bar(stat = "identity") +
  # clean up x-axis 
  theme(axis.title.x = element_blank(), #remove x-axis label 
        axis.text.x =  element_text(angle = 90, #rotate tick labels 
                                    vjust = 0.5,
                                    hjust = 0.5,
                                    size = 8)) +
  # clean up y-axis
  ylab("Relative Abundance\n(% total sequences)") +
  ggtitle("Week 3 - Negative Butyrate Status")+
  scale_fill_manual(name = "Phylum",
                  #these are all the phylum categories
                  breaks = c("Actinobacteria", "Bacteria_unclassified", 
                             "Bacteroidetes", "Elusimicrobia", 
                             "Firmicutes", "Fusobacteria", 
                             "Lentisphaerae", "Proteobacteria", 
                             "Sprichaetes", "Synergistetes", 
                             "Verrucomicrobia"), 
                  #these are updated names for the phyla
                  labels = c("Actinobacteria", "Unclassified", 
                             "Bacteroidetes", "Elusimicrobia", 
                             "Firmicutes", "Fusobacteria", 
                             "Lentisphaerae", "Proteobacteria", 
                             "Sprichaetes", "Synergistetes", 
                             "Verrucomicrobia"), 
                  #colors for each category
                  values = c("#4bb092", "#618bcb",
                             "#b95d6a", "#79853d",
                             "#c783c1", "#bd7a43",
                             "#c9aa3c", "#d4447a",
                             "#7663cb", "#67b34a",
                             "#d35238")  )
negbut_plotwk3
```

##Plot for Positive Status Week 1
```{r}
posbut_plotwk1 <- physq_phyla %>%
  # filter for supplement and study week 
  filter(butyrate_status == "positive",
         study_week == "week1") %>% 
  # set parameters for plot
  ggplot(aes(x = participant_id, y = relative_abundance,
             fill = Phylum,
             study_week, frequency)) + 
  # layer for stacked bar plot 
  geom_bar(stat = "identity") +
  # clean up x-axis 
  theme(axis.title.x = element_blank(), #remove x-axis label 
        axis.text.x =  element_text(angle = 90, #rotate tick labels 
                                    vjust = 0.5,
                                    hjust = 0.5,
                                    size = 8)) +
  # clean up y-axis
  ylab("Relative Abundance\n(% total sequences)") +
  ggtitle("Week 1 - Positive Butyrate Status")+
  scale_fill_manual(name = "Phylum",
                  #these are all the phylum categories
                  breaks = c("Actinobacteria", "Bacteria_unclassified", 
                             "Bacteroidetes", "Elusimicrobia", 
                             "Firmicutes", "Fusobacteria", 
                             "Lentisphaerae", "Proteobacteria", 
                             "Sprichaetes", "Synergistetes", 
                             "Verrucomicrobia"), 
                  #these are updated names for the phyla
                  labels = c("Actinobacteria", "Unclassified", 
                             "Bacteroidetes", "Elusimicrobia", 
                             "Firmicutes", "Fusobacteria", 
                             "Lentisphaerae", "Proteobacteria", 
                             "Sprichaetes", "Synergistetes", 
                             "Verrucomicrobia"), 
                  #colors for each category
                  values = c("#4bb092", "#618bcb",
                             "#b95d6a", "#79853d",
                             "#c783c1", "#bd7a43",
                             "#c9aa3c", "#d4447a",
                             "#7663cb", "#67b34a",
                             "#d35238")  )
posbut_plotwk1
```

##Plot for Positive Status Week 3
```{r}
posbut_plotwk3 <- physq_phyla %>%
  # filter for supplement and study week 
  filter(butyrate_status == "positive",
         study_week == "week3") %>% 
  # set parameters for plot
  ggplot(aes(x = participant_id, y = relative_abundance,
             fill = Phylum,
             study_week, frequency)) + 
  # layer for stacked bar plot 
  geom_bar(stat = "identity") +
  # clean up x-axis 
  theme(axis.title.x = element_blank(), #remove x-axis label 
        axis.text.x =  element_text(angle = 90, #rotate tick labels 
                                    vjust = 0.5,
                                    hjust = 0.5,
                                    size = 8)) +
  # clean up y-axis
  ylab("Relative Abundance\n(% total sequences)") +
  ggtitle("Week 3 - Positive Butyrate Status")+
  scale_fill_manual(name = "Phylum",
                  #these are all the phylum categories
                  breaks = c("Actinobacteria", "Bacteria_unclassified", 
                             "Bacteroidetes", "Elusimicrobia", 
                             "Firmicutes", "Fusobacteria", 
                             "Lentisphaerae", "Proteobacteria", 
                             "Sprichaetes", "Synergistetes", 
                             "Verrucomicrobia"), 
                  #these are updated names for the phyla
                  labels = c("Actinobacteria", "Unclassified", 
                             "Bacteroidetes", "Elusimicrobia", 
                             "Firmicutes", "Fusobacteria", 
                             "Lentisphaerae", "Proteobacteria", 
                             "Sprichaetes", "Synergistetes", 
                             "Verrucomicrobia"), 
                  #colors for each category
                  values = c("#4bb092", "#618bcb",
                             "#b95d6a", "#79853d",
                             "#c783c1", "#bd7a43",
                             "#c9aa3c", "#d4447a",
                             "#7663cb", "#67b34a",
                             "#d35238")  )
posbut_plotwk3
```

##Combination Plot
```{r}
negbut_comb_plot <- plot_grid(negbut_plotwk1, negbut_plotwk3, posbut_plotwk1, posbut_plotwk3, align = "h", nrow = 2, ncol = 2)

negbut_comb_plot


# save plot
save_plot(negbut_comb_plot, 
          nrow = 2 , ncol = 2, 
          filename = "~/Documents/UMBio201/Final Project/Projects/negbut_ra_comb_plot.pdf")


```


#Difference in Richness between positive and negative respondents

```{r}
# calculate richness
richness_df <- physq_but %>%
  # calculate richness
  estimate_richness(., split = TRUE,  measures = c("Observed")) %>% 
  # make sample_id column before join 
  rownames_to_column(var = "id") %>% 
  # join with sample data imported above
  inner_join(butyrate_df, by = "id") %>%  
  rename(richness = Observed) %>%
  # calculate average richness per person, per week 
  group_by(participant_id, study_week, semester.x, 
          butyrate_status) %>%
  summarise(avg_richness = round(mean(richness), digits = 0)) 

richness_df
dim(richness_df)
```

## Richness plot(s)
```{r}
rich1 <- physq_but %>%
  plot_richness(., "study_week", measures = c("Observed")) +
  facet_grid("butyrate_status") +
  ylab("Richness (Observed ESVs)") + xlab(NULL)
rich1

rich1$layers <- rich1$layers[-1] #drop geom_point 

rich2 <- rich1 +   
  geom_violin(aes(color = study_week)) + #add violin in color
  geom_jitter(aes(color = study_week)) +  #add individual points in color 
  theme(legend.position = "none")
rich2

# save plots
save_plot(rich1, 
          nrow = 2 , ncol = 2, 
          filename = "~/Documents/UMBio201/Final Project/Projects/richness_1.pdf")

# save plot
save_plot(rich2, 
          nrow = 2 , ncol = 2, 
          filename = "~/Documents/UMBio201/Final Project/Projects/rich2.pdf")
```


```{r}


#Sample Sizes
richness_df%>%
  group_by(butyrate_status) %>%
  summarise(sample_size = n())


```

```{r}
# statistical test(s)

repos_wk1 <- richness_df %>%
  filter(butyrate_status == "positive",
         study_week == "week1")
  shapiro.test(repos_wk1$avg_richness) #p-value = 0.2289
  
repos_wk3 <- richness_df %>%
  filter(butyrate_status == "positive",
         study_week == "week3")
  shapiro.test(repos_wk3$avg_richness) #p-value = 0.6196
  
reneg_wk1 <- richness_df %>%
  filter(butyrate_status == "negative",
         study_week == "week1")
  shapiro.test(reneg_wk1$avg_richness) #p-value = 0.8638
  
reneg_wk3 <- richness_df %>%
  filter(butyrate_status == "negative",
         study_week == "week3")
  shapiro.test(reneg_wk3$avg_richness) #p-value = 0.512
  
ggplot(repos_wk1, aes(x=avg_richness)) +
  geom_histogram() 
qqnorm(repos_wk1$avg_richness); qqline(repos_wk1$avg_richness)

ggplot(reneg_wk1, aes(x=avg_richness)) +
  geom_histogram() 
qqnorm(reneg_wk1$avg_richness); qqline(reneg_wk1$avg_richness)

ggplot(repos_wk3, aes(x=avg_richness)) +
  geom_histogram() 
qqnorm(repos_wk3$avg_richness); qqline(repos_wk3$avg_richness)

ggplot(reneg_wk1, aes(x=avg_richness)) +
  geom_histogram() 
qqnorm(reneg_wk3$avg_richness); qqline(reneg_wk3$avg_richness)

#check variances
  var.test(x = repos_wk1$avg_richness, 
         y = repos_wk3$avg_richness, 
         alternative = "two.sided") 
  #p-value = 0.01683, variances are not equal
   
  var.test(x = reneg_wk1$avg_richness, 
         y = reneg_wk3$avg_richness, 
         alternative = "two.sided") 
#p-value= 0.08123, variances are equal

```

```{r}
#Statistical Test
t.test(x = repos_wk1$avg_richness, 
       y = repos_wk3$avg_richness,  
       var.equal = FALSE,
       paired = TRUE,
       alternative = "two.sided")
#p-value -  0.06531

t.test(x = reneg_wk1$avg_richness, 
       y = reneg_wk3$avg_richness,  
       var.equal = TRUE,
       paired = TRUE,
       alternative = "two.sided")

#p-value: 8.909e-06

```

Conclusion: YES there was a highly significant change in richness for the negative responses from week 1 to week 3 (#p-value: 8.909e-06), and a marginally significant (only when using 0.10 significance threshold, p-value = 0.06531) for the positive response individuals from week1 to week3.


## Community Composition

Conduct an ordination using the Bray-Curtis index to determine if the communities of participants were different based on whether they had a positve or negative change in butyrate.

### Week 1
```{r}
# data formatting 
physq_but1 <- physq_but %>% 
  subset_samples(., study_week == "week1") %>% #composition before the consumption of starch
  prune_taxa(taxa_sums(.) > 1000, .) %>%
  prune_samples(sample_sums(.) > 1000, .)

# get read counts 
but_sum_df <- data.frame(sum = sample_sums(physq_but1))

# Histogram of sample read counts
ggplot(but_sum_df, aes(x = sum)) + 
  geom_histogram(color = "black", fill = "gray", binwidth = 2500) +
  ggtitle("Distribution of sample sequencing depth") + 
  xlab("Read counts") +
  theme(axis.title.y = element_blank())

# Summary statistics on read counts 
min(sample_sums(physq_but1)) #6271
mean(sample_sums(physq_but1)) #19450.17
max(sample_sums(physq_but1))  #34432

# scale samples to even depth using custom function
physq_scale <- physq_but1 %>%
  scale_reads(round = "round") 

#you want the stress to be less than 0.20
```

```{r}
# reads coutns + subsample
#see above
```

```{r}
# ordination

physq_butbc_1 <- ordinate(physq_scale, 
           method = "NMDS", 
           k=3, maxit=500, try=50,
           distance = "bray")
physq_butbc_1

#you want the stress to be less than 0.20
```

```{r}
# plot

ordplot_but <- plot_ordination(physeq = physq_but1, 
                     ordination = physq_butbc_1, 
                     type = "samples", 
                     color = "butyrate_status")
print(ordplot_but)

# save plot
save_plot(ordplot_but, 
          nrow = 2 , ncol = 2, 
          filename = "~/Documents/UMBio201/Final Project/Projects/ordination_but.pdf")

#Find outlier individuals (super positive/super negative -- look at dietary info for shared characteristics (fiber and protein))
```

```{r}
# Bray-Curtis 

# calculate BC index, get distance matrix
dat_bray_1 <- phyloseq::distance(physq_but1, method = "bray") 

samplebdf <- physq_but1%>% 
  sample_data(.) %>% #extract sample data from phyloseq object 
  as(., "data.frame") #convert to data frame for adonis()

# run test
adn_res <- adonis(formula = dat_bray_1 ~ butyrate_status, 
                  data = samplebdf)

# view results 
print(adn_res)

# Jaccard 

# calculate index, get distance matrix
dat_jac_1 <- phyloseq::distance(physq_but1, method = "jaccard") 

samplejdf <- physq_but1%>% 
  sample_data(.) %>% #extract sample data from phyloseq object 
  as(., "data.frame") #convert to data frame for adonis()

# run test
adn_res2 <- adonis(formula = dat_jac_1 ~ butyrate_status, 
                  data = samplejdf)

# view results 
print(adn_res2)
```
Conclusion: High p-value and low R2 for both (Bray: R2 = 0.00632, p-value= 0.623, Jaccard, R2 = 0.00664,  p-value = 0.626) indicate that there may be no significant initial difference in community composition between the postive and negative respondents. 

### Week 3
```{r}
# data formatting 
physq_but2 <- physq_but %>% 
  subset_samples(., study_week == "week3") %>% #composition after the consumption of starch
  prune_taxa(taxa_sums(.) > 1000, .) %>%
  prune_samples(sample_sums(.) > 1000, .)

# get read counts 
but_sum_df2 <- data.frame(sum = sample_sums(physq_but2))

# Histogram of sample read counts
ggplot(but_sum_df2, aes(x = sum)) + 
  geom_histogram(color = "black", fill = "gray", binwidth = 2500) +
  ggtitle("Distribution of sample sequencing depth") + 
  xlab("Read counts") +
  theme(axis.title.y = element_blank())

# Summary statistics on read counts 
min(sample_sums(physq_but2)) #8557
mean(sample_sums(physq_but2)) #20379.15
max(sample_sums(physq_but2))  #43631

# scale samples to even depth using custom function
physq_scale2 <- physq_but2 %>%
  scale_reads(round = "round") 

#you want the stress to be less than 0.20
```

```{r}
# reads coutns + subsample
#see above
```

```{r}
# ordination

physq_butbc_2 <- ordinate(physq_scale2, 
           method = "NMDS", 
           k=3, maxit=500, try=50,
           distance = "bray")
physq_butbc_2

#you want the stress to be less than 0.20
```

```{r}
# plot

ordplot_but2 <- plot_ordination(physeq = physq_but2, 
                     ordination = physq_butbc_2, 
                     type = "samples", 
                     color = "butyrate_status")
print(ordplot_but2)

#Find outlier individuals (super positive/super negative -- look at dietary info for shared characteristics (fiber and protein))

# save plot
save_plot(ordplot_but2, 
          nrow = 2 , ncol = 2, 
          filename = "~/Documents/UMBio201/Final Project/Projects/ordination_but2.pdf")
```

```{r}
# Bray-Curtis 

# calculate BC index, get distance matrix
dat_bray_2 <- phyloseq::distance(physq_but2, method = "bray") 

samplebdf2 <- physq_but2%>% 
  sample_data(.) %>% #extract sample data from phyloseq object 
  as(., "data.frame") #convert to data frame for adonis()

# run test
adn_res3 <- adonis(formula = dat_bray_2 ~ butyrate_status, 
                  data = samplebdf2)

# view results 
print(adn_res3)

# Jaccard 

# calculate index, get distance matrix
dat_jac_2 <- phyloseq::distance(physq_but2, method = "jaccard") 

samplejdf2 <- physq_but2%>% 
  sample_data(.) %>% #extract sample data from phyloseq object 
  as(., "data.frame") #convert to data frame for adonis()

# run test
adn_res4 <- adonis(formula = dat_jac_2 ~ butyrate_status, 
                  data = samplejdf2)

# view results 
print(adn_res4)
```
Conclusion: High p-value and low R2 for both (Bray: R2 = 0. 0.00807, p-value=  0.282, Jaccard, R2 = 0.00765,  p-value =  0.286)indicate that there may be no significant final difference in community composition between the postive and negative respondents. 



#Summary Stats for Butyrate Delta

```{r}
#Summary Statistics for Butyrate Delta

summary(butyrate_delta$delta_but)


```
| |Min | Median| Mean |Max |
|:-----:|:-----:|:-----:|:-----:|
| -20.830 | 1.000  | 1.419 |30.210  |


#Look at the Two Outlier Individuals
```{r}
#Make a Df with just the two individuals 

outlierdf <- butyrate_delta %>%
  filter(delta_but >= 30 | delta_but <= -20)

#Very Positive: U132
#Very Negative: U778

```




#Antibiotic Correlations


##Import Antibiotic Data
```{r}

antibioticdf <- read_delim("antibiotic_data.txt", 
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE) %>%
  # make all column names lower case
  rename_all(tolower) %>%
  distinct(participant_id, .keep_all = TRUE)

butantibdf <- merge(butyrate_df, antibioticdf, by = "participant_id", all.x= TRUE) %>%
   select(-id) %>%
  filter(antibiotic_past_year != "unknown") %>%
    spread(study_week, butyrate_mean)  %>%
  # drop NAs from select columns
  drop_na(week1, week3)  %>% 
  gather(study_week, butyrate_mean, week1, week3) 
```

##Proportion Table/Test for Antiobiotic Status and Butyrate Status 
```{r}

#Proportion Test

butantiprop_tab <- with(butantibdf, table(antibiotic_past_year, butyrate_status))

butantiprop_tab

prop.test(butantiprop_tab, 
          alternative = "two.sided",
          correct = FALSE)
```
Conclusion: p-value of 0.1174 idnicates no significant difference in proportions of negative/positive butyrate status by antibiotic status.

##T-Test for Antibiotic Status and Delta Butyrate
```{r}
#Unpaired Two Sample T-test

anti2_yes_df <- butantibdf%>%
  filter(antibiotic_past_year == "Yes" ) %>%
  rename(anti_but_yes = "delta_but" ) %>%
  select(-study_week)

anti2_no_df <- butantibdf %>%
  filter(antibiotic_past_year == "No" ) %>%
  rename(anti_but_no = "delta_but") %>%
  select(-study_week)

antibut_plot <- butantibdf %>%
  ggplot(aes(x = study_week, y = delta_but)) +
  geom_violin(aes(color = study_week)) + geom_jitter(aes(color = study_week)) +
  facet_grid(~antibiotic_past_year) +
  xlab("Study Week") +
  ylab("Butyrate Mean Concentration")+
  theme(legend.position = "none")
antibut_plot


#Sample Sizes

summarise(anti2_yes_df, sample_size = n()) #134	

summarise(anti2_no_df, sample_size = n()) #202

#Normality Testing
shapiro.test(anti2_yes_df$anti_but_yes) #p-value = 0.2536 -- roughly normal distribution by CLT

  ggplot(anti2_yes_df, aes(x = anti_but_yes)) + geom_histogram()

  qqnorm(anti2_yes_df$anti_but_yes); qqline(anti2_yes_df$anti_but_yes)

shapiro.test(anti2_no_df$anti_but_no) #p-value = 0.02084



#Variance testing

var.test(x = anti2_yes_df$anti_but_yes, 
         y = anti2_no_df$anti_but_no, 
         alternative = "two.sided") 
#p-value = 0.2385, variances equal


#Antibiotic table

table(butantibdf$antibiotic_past_year)  
    
#t-test for antibiotic status and value of delta butyrate 
t.test(x = anti2_yes_df$anti_but_yes, 
         y = anti2_no_df$anti_but_no,
       var.equal = TRUE, 
       alternative = "less")
#p-calue = 0.3195

```
Conclusion: p-value of 0.3195 indicates no significant difference in delta butyrate across the two types of antibiotic statuses ("Yes" or "No")


##Outlier Antibiotics
```{r}
outlierantibdf <- merge(outlierdf, antibioticdf, by = "participant_id", all.x= TRUE)

#Negative individual did not consume antibiotics, Positive individual is unknown

```

#Look at Dietary Commonalities


##Import and Merge 
```{r}
veg_meatdf <- read_delim("veg_meat_data.txt", 
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE) %>%
  filter(diet_status == "Omnivorous" |
        diet_status == "Vegetarian" |
        diet_status == "Pescatarian" |
        diet_status == "Vegan" |
        diet_status == "Flexitarian" |
        diet_status == "Non-red Omnivorous") %>%
  # make all column names lower case
  rename_all(tolower) %>%
  distinct(participant_id, .keep_all = TRUE)

butvegmeatdf <- merge(butyrate_df, veg_meatdf, by = "participant_id", all.x= TRUE) %>%
   select(-id) %>%
    spread(study_week, butyrate_mean)  %>%
  # drop NAs from select columns
  drop_na(week1, week3) %>% 
  gather(study_week, butyrate_mean, week1, week3) %>%
  mutate(id = paste(participant_id, study_week, sep = "_")) %>%
  drop_na 

```

##Proportion Test for Butyrate Status and Diet Status 
```{r}

butvegmeatprop_df<- butvegmeatdf %>%
    filter(diet_status == "Omnivorous" |
        diet_status == "Vegetarian" ) 

butvegmeatprop_df %>%
  group_by(diet_status, butyrate_status) %>%
  summarise(Counts = n())

butvegmeatprop_tab <- with(butvegmeatprop_df, table(diet_status, butyrate_status))

butvegmeatprop_tab

prop.test(butvegmeatprop_tab, 
          alternative = "two.sided",
          correct = FALSE)
#p-value = 0.3756
```
 Conclusion: The p-value of 0.3756 for this proportion test indicates that the proportion of butyrate status responses are not different for either type of diet.

##Kruskal-Wallace Test for Diet Status and Delta Butyrate
```{r}

butvegmeatdf %>%
  ggplot(aes(x = diet_status,
             y = delta_but,
             color = diet_status)) +
# geom_hline adds a horitzonal line which makes it easy to quickly see which supplements had a positive or negative influence on the concentration
  geom_hline(yintercept = 0,
             linetype = "dashed") + 
  geom_violin() + #add violion plots over top the hline annotation
  geom_jitter() + # add jitter points overtop of the violin 
  xlab(NULL) + #hide x-axis label 
  ylab("Delta Butyrate Concentration") + 
  theme(legend.position = "none", #hide legend, not adding anything we can't determine from axis labels
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) #rotate x-axis tick mark labels

#Sample Sizes

butvegmeatdf %>%
  group_by(diet_status) %>%
  summarise(sample_size = n())

# check normality of each group 
s1 <- butvegmeatdf %>%
  filter(diet_status == "Vegetarian") 
shapiro.test(s1$delta_but) #p-value = 9.423e-07

s2 <- butvegmeatdf %>%
  filter(diet_status == "Non-red Omnivorous") 
shapiro.test(s2$delta_but) #p-value = 0.01084

s3 <- butvegmeatdf %>%
  filter(diet_status == "Omnivorous") 
shapiro.test(s3$delta_but) #p-value = 0.009345

s4 <- butvegmeatdf %>%
  filter(diet_status == "Vegan") 
shapiro.test(s4$delta_but) #p-value = 0.00145

s5 <- butvegmeatdf %>%
  filter(diet_status == "Pescatarian") 
shapiro.test(s5$delta_but) #p-value =  0.02083

#check variances
bartlett.test(delta_but ~ diet_status, data = butvegmeatdf)
##not equal

#Kruskal-Wallace Test

kruskal.test(delta_but ~ diet_status, data = butvegmeatdf)

#p-value = 0.6902

```
Conclusion: P-value of 0.6902 from non-parametric Kruskall-Wallace test indicates no significant difference in delta butyrate across the 5 groups.

##Save Dietary Curated Data Frame

```{r}
# save final data frame
write_delim(veg_meatdf , path = "~/Documents/UMBio201/Projects/vegmeat_curated_data.txt", delim = "\t")

```
